---
title: "Summary Statistical Inference Course"
author: "Desiré De Waele"
date: "24 maart 2016"
output:
  html_document:
    fig_height: 4
    fig_width: 5
    highlight: tango
    theme: united
    toc: yes
    toc_depth: 2
---

#Conditional probability
##Definition
$$ P(A ~|~ B) = \frac{P(A \cap B)}{P(B)} $$

##Bayes' rule

$$ P(B ~|~ A) = \frac{P(A ~|~ B) P(B)}{P(A ~|~ B) P(B) + P(A ~|~ B^c)P(B^c)}. $$

##Diagnostic tests

* Sensitivity - probability that the test is positive given that the subject actually has the disease, $P(+ ~|~ D)$
* Specificity - probability that the test is negative given that the subject does not have the disease, $P(- ~|~ D^c)$
* Positive predicted value - probability that the subject has the disease given that the test is positive, $P(D ~|~ +)$
* Negative predicted value - probability that the subject does not have the disease given that the test is negative, $P(D^c ~|~ -)$
* Prevalence - marginal probability of disease, $P(D)$

#Expected values
##Population Mean
$$ E[X] = \sum_x xp(x)$$

##Sample Mean
$$ \bar X = \sum_{i=1}^n x_i p(x_i)$$

```{r, warning=FALSE, message=FALSE, fig.width=10}
library(UsingR); data(galton); library(ggplot2); library(reshape2)
longGalton <- melt(galton, measure.vars = c("child", "parent"))
g <- ggplot(longGalton, aes(x = value))
g <- g + geom_histogram(aes(y = ..density.., fill = variable), binwidth=1, color = "black")
g <- g + geom_density(size = 1)
g <- g + facet_grid(. ~ variable)
g
```

#Variance

## Population Variance
$$ \sigma^2 = Var(X) = E[(X - \mu)^2] = E[X^2] - E[X]^2 $$

## Sample Variance
$$ S^2 = \frac{\sum_{i=1} (X_i - \bar X)^2}{n-1} $$

## Standard Error of the Mean
$$ S / \sqrt{n} $$

#Common distributions
##Bernoulli distribution
$$ P(X = x) = \left( \begin{array}{c} n \ x \end{array} \right) p^x(1 - p)^{n-x} $$
Suppose a friend has 8 children, 7 of which are girls and none are twins. If each gender has an independent 50% probability for each birth, what's the probability of getting 7 or more girls out of 8 births? 
$$\left( \begin{array}{c} 8 \ 7 \end{array} \right) .5^{7}(1-.5)^{1} + \left( \begin{array}{c} 8 \ 8 \end{array} \right) .5^{8}(1-.5)^{0} \approx 0.04 $$

##Normal distribution

* Approximately 68\%, 95\% and 99\% of the normal density lies within 1, 2 and 3 standard
deviations from the mean, respectively.
* 1.28, 1.645, 1.96 and 2.33 are the 90th, 95th, 97:5th and 99th percentiles of the
standard normal distribution, respectively.

##Poisson distribution

#Central Limit Theorem
The sample means of any distribution approximate a normal distribution. The mean is the population mean $\mu$, the variance is the standard error, the population variance $\sigma$ divided by $\sqrt{n}$.
$$\frac{\bar X_n - \mu}{\sigma / \sqrt{n}}= \frac{\sqrt n (\bar X_n - \mu)}{\sigma} = \frac{\mbox{Estimate} - \mbox{Mean of estimate}}{\mbox{Std. Err. of estimate}}$$
```{r, fig.width=10}
nosim <- 1000
cfunc <- function(x, n) sqrt(n) * (mean(x) - 3.5) / 1.71
dat <- data.frame(x = c(apply(matrix(sample(1 : 6, nosim * 10, replace = TRUE), nosim), 1, cfunc, 10),
                        apply(matrix(sample(1 : 6, nosim * 20, replace = TRUE), nosim), 1, cfunc, 20),
                        apply(matrix(sample(1 : 6, nosim * 30, replace = TRUE), nosim), 1, cfunc, 30)),
                  size = factor(rep(c(10, 20, 30), rep(nosim, 3))))

g <- ggplot(dat, aes(x = x, fill = size))
g <- g + geom_histogram(alpha = .50, binwidth=.3, colour = "black", aes(y = ..density..)) 
g <- g + stat_function(fun = dnorm, size = 1)
g <- g + facet_grid(. ~ size); g
```

